The Iterative Signature Algorithm for Gene Expression data

### 2009-08-05 eisa/inst/doc/EISA_tutorial.Rnw
EISA vignette improvements.

% TODO: length, dim, featureNames, sampleNames, pData, 
% annotation, organism, indexing

The \Rfunction{getNoFeatures} function returns a numeric vector, the
number of features (probesets in our case) in each module. Similarly,
\Rfunction{getNoSamples} returns a numeric vector, the number of
samples in each module. 
<<number of features and samples>>=
getNoFeatures(modules)
getNoSamples(modules)
@

@@

$\mu$RNAs are short RNA molecules that regulate gene expression.
TargetScan is a data based of predicted target genes of $\mu$RNAs, for
severral organisms. There are two R packages that incorporate this
database, one for human and another one for mouse, right now they can
be downloaded from
\url{http://www2.unil.ch/cbg/index.php?title=Building_BioConductor_Annotation_Packages}.
The \Rpackage{targetscan.Hs.eg.db} package is for human, the
\Rpackage{targetscan.Mm.eg.db} package is for mouse.

@@

You can also use the \Rpackage{biclust} package to create image
plots, by calling the \Rfunction{drawHeatmap} function. The result is
in Fig.~\ref{fig:drawHeatmap}.
<<heatmap biclust pre,echo=FALSE,results=hide>>=
CairoPNG("heatmap.png", 400, 400 )
@ 
<<heatmap>>=
drawHeatmap(exprs(ALL[featureNames(modules),]), Bc, number=1) 
@
<<heatmap post,echo=FALSE,results=hide>>=
dev.off() 
@ 
\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{heatmap}
  \caption{TODO}
  \label{fig:drawHeatmap}
\end{figure}

@@

>>>>

Calling \Rfunction{length} on \variable{modules} returns the number of
ISA modules in the set, \Rfunction{dim} gives the dimension of the
input expression matrix: the number of features and the number of
samples.
<<accessors 1>>=
length(modules)
dim(modules)
@

Functions \Rfunction{featureNames} and \Rfunction{sampleNames} return
the names of the features and samples, just like the functions with
the same name for an \Rclass{ExpressionSet}
<<accessors 2>>=
featureNames(modules)[1:5]
sampleNames(modules)[1:5]
@

The \Rfunction{getNoFeatures} function returns a numeric vector, the
number of features (probesets in our case) in each module. Similarly,
\Rfunction{getNoSamples} returns a numeric vector, the number of
samples in each module. Similarly, \Rfunction{pData} returns the
phenotype data of the expression set as a data frame. The
\Rfunction{getOrganism} function returns the scientific name of the
organism, \Rfunction{annotation} the name of the chip. For the former
the appropriate annotation must be installed.
<<number of features and samples>>=
getNoFeatures(modules)
getNoSamples(modules)
colnames(pData(modules))
getOrganism(modules)
annotation(modules)
@

The double bracket indexing operator (`\verb+[[+') can be used to
select some modules from the complete set, the result is another,
smaller \Rclass{ISAModules} object. The following selects the first
five modules.
<<indexing>>=
modules[[1:5]]
@

The single bracket indexing operator can be used to restrict an
\Rclass{ISAModules} object to a subset of features and/or
samples. E.g. selecting all features that map to a gene on
chromosome~1 can be done with 
<<indexing 2>>=
chr <- get(paste(annotation(modules), sep="", "CHR"))
entrez <- sapply(mget(featureNames(modules), chr), 
                 function(x) "1" %in% x)
modules[entrez,]
@

Similarly, selecting all B-cell samples can be performed with
<<indexing 3>>=
modules[ ,grep("^B", pData(modules)$BT)]
@

@@

$\mu$RNAs are short RNA molecules that regulate gene expression.
TargetScan is a data based of predicted target genes of $\mu$RNAs, for
several organisms. There are two R packages that incorporate this
database, one for human and another one for mouse, right now they can
be downloaded from
\url{http://www2.unil.ch/cbg/index.php?title=Building_BioConductor_Annotation_Packages}.
The \Rpackage{targetscan.Hs.eg.db} package is for human, the
\Rpackage{targetscan.Mm.eg.db} package is for mouse.

@@

You can also use the \Rpackage{biclust} package to create image
plots, by calling the \Rfunction{drawHeatmap} function. The result is
in Fig.~\ref{fig:drawHeatmap}.
<<heatmap biclust pre,echo=FALSE,results=hide>>=
CairoPNG("heatmap.png", 400, 400 )
@ 
<<heatmap>>=
drawHeatmap(exprs(ALL[featureNames(modules),]), Bc, number=1) 
@
<<heatmap post,echo=FALSE,results=hide>>=
dev.off() 
@ 
\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{heatmap}
  \caption{Heatmap of the first transcription module.}
  \label{fig:drawHeatmap}
\end{figure}

@@

The \Rfunction{ISA} function takes care of all steps performed during
a modular study, and for each step it uses parameters work reasonably
well. In some cases, however, one wants to access these steps
individually, to use custom parameters instead of the defaults.

In this section, we will still use the acute lymphoblastic leukemia
gene expression data from the \Rpackage{ALL} package.

\subsection{Non-specific filtering}

The first step of the analysis typically involves non-specific
filtering of the probesets. The aim is to eliminate the probesets that
do not show variation across the samples, as they only contribute
noise to the data.

By default (i.e. if the \Rfunction{ISA} function is called) this is
performed using the \Rpackage{genefilter} package, and the default
filter is based on the inter-quantile ratio of the probesets'
expression values, a robust measure of variance. Other possible
filtering techniques include using the AffyMetrix present/absent calls
produced by the \Rfunction{mas5calls} function of the \Rpackage{affy}
package, but this requires the raw data, so in this vignette we use a
simple method based on variance and minimum expression value: only
probesets that have a variance of at least \variable{varLimit} and
that have at least \variable{kLimit} samples with expression values
over \variable{ALimit} are kept.
<<filtering>>=
varLimit <- 0.5
kLimit <- 4
ALimit <- 5
flist <- filterfun(function(x) var(x)>varLimit, kOverA(kLimit,ALimit))
ALL.filt <- ALL[genefilter(ALL, flist), ]
@

The original expression set had \Sexpr{nrow(ALL)} features, the
filtered one has only \Sexpr{nrow(ALL.filt)}.

\subsection{Entrez Id matching}

In this step we match the probesets to Entrez Id and remove the ones
that don't map to any Entrez genes.
<<Entrez>>=
ann <- annotation(ALL.filt)
library(paste(ann, sep=".", "db"), character.only=TRUE)
ENTREZ <- get( paste(ann, sep="", "ENTREZID") )
EntrezIds <- mget(featureNames(ALL.filt), ENTREZ)
keep <- sapply(EntrezIds, function(x) length(x) >= 1 && !is.na(x))
ALL.filt.2 <- ALL.filt[keep,]
@

To reduce ambiguity in the interpretation of the results, we might
also want to keep only single probeset for each Entrez genes.
<<Entrez unique>>=
vari <- apply(exprs(ALL.filt.2), 1, var)
larg <- findLargest(featureNames(ALL.filt.2), vari, data=annotation(ALL.filt.2))
ALL.filt.3 <- ALL.filt.2[larg,]
@

\subsection{Normalizing the data}

The ISA works best, if the expression matrix is scaled and
centered. In fact, the two sub-steps of an ISA step require expression
matrices that are normalized differently. The
\Rfunction{ISA.normalize} function can be used to calculate the
normalized expression matrices; it returns an \Rclass{ExpressionSet}
object including three expression matrices: the original raw
expression, the row-wise normalized and the column-wise normalized
expression matrix.
<<normed>>=
ALL.normed <- ISA.normalize(ALL.filt.3)
ls(assayData(ALL.normed))
@

\subsection{Generating starting seeds for the ISA}

The ISA is an iterative algorithm that starts with a set of input
seeds. An input seed is basically a set of probesets and the ISA stepwise
refines this set by 1) including other probesets in the set that are
coexpressed with the input probesets and 2) removing probesets from it
that are not coexpressed with the rest of the input set.

The \Rfunction{generate.seeds} generates a set of random seeds (i.e. a
set of random gene sets). See its documentation if you need to change
the sparsity of the seeds.

<<seeds>>=
random.seeds <- generate.seeds(length=nrow(ALL.normed), count=100)
@

In addition to random seeds, it is possible to start the ISA iteration
from ``educated'' seeds, i.e. gene sets the user is interested in, or
a set of samples that are supposed to have coexpressed genes. We
create another set of starting seeds here, based on the type of acute
lymphoblastic leukemia: ``B'', ``B1'', ``B2'', ``B3'', ``B4'' or
``T'', ``T1'', ``T2'', ``T3'' and ``T4''.
<<smart seeds>>=
type <- as.character(pData(ALL.normed)$BT)
ss1 <- ifelse(grepl("^B", type), -1, 1)
ss2 <- ifelse(grepl("^B1", type), 1, 0)
ss3 <- ifelse(grepl("^B2", type), 1, 0)
ss4 <- ifelse(grepl("^B3", type), 1, 0)
ss5 <- ifelse(grepl("^B4", type), 1, 0)
ss6 <- ifelse(grepl("^T1", type), 1, 0)
ss7 <- ifelse(grepl("^T2", type), 1, 0)
ss8 <- ifelse(grepl("^T3", type), 1, 0)
ss9 <- ifelse(grepl("^T4", type), 1, 0)
smart.seeds <- cbind(ss1, ss2, ss3, ss4, ss5, ss6, ss7, ss8, ss9)
@

The \variable{ss1} seed includes all samples, but their sign is
opposite for B-cell leukemia samples and T-cell samples. This way ISA
is looking for sets of genes that are differently regulated in these
two groups of samples. \variable{ss2} contains only B1 type samples,
so here we look for genes that are specific to this variant of the
disease. The other seeds are similar, for the other subtypes.

\subsection{Performing the ISA iteration}

We perform the ISA iterations for our two sets of seeds separately.
<<iteration>>=
modules1 <- ISA.iterate(ALL.normed, row.seeds=random.seeds, 
                        thr.row=2, thr.col=2)
modules2 <- ISA.iterate(ALL.normed, col.seeds=smart.seeds,
                        thr.row=2, thr.col=2)
@

\subsection{Dropping non-unique modules}

\Rfunction{ISA.iterate} returns the same number of ``modules'' as the
number of input seeds; these, however, do not always correspond to
meaningful modules, the input seeds can converge to an all-zero
vector, or occasionally they may not converge at all. It is also
possible that two or more input seeds converge to the same module.

The \Rfunction{ISA.unique} function eliminates the all-zero or
non-convergent input seeds and keeps only one instance of the
duplicated ones.
<<unique>>=
modules1.unique <- ISA.unique(ALL.normed, modules1)
modules2.unique <- ISA.unique(ALL.normed, modules2)
@

\Sexpr{length(modules1.unique)} modules were kept for the first set of
seeds and \Sexpr{length(modules2.unique)} for the second set.

\subsection{Dropping non-robust modules}

The \Rfunction{ISA.filter.robust} function filters a set of modules by
running ISA with the same parameters on a scrambled data set and then
calculating a robustness score, both for the real modules and the ones
from the scrambled data. The highest robustness score obtained from
the scrambled data is used as a threshold to filter the real modules.
<<robust>>=
modules1.robust <- ISA.filter.robust(ALL.normed, modules1.unique)
modules2.robust <- ISA.filter.robust(ALL.normed, modules2.unique)
@

We still have \Sexpr{length(modules1.robust)} modules for
the first set of seeds and \Sexpr{length(modules2.unique)} for the
second set.

\subsection{Differentially regulated modules}

Now we check whether the ISA modules that we found can be used as
classifiers for the different types of ALL. For this we use the sample
scores of the modules, the score of a sample in a given module is the
(weighted) average of the expression of the genes of the module in
the sample in question.

Let's first check, whether any of the modules can distinguish between
T-cell and B-cell ALL samples.
<<B vs T>>=
scores1 <- getSampleMatrix(modules1.robust)
tt1 <- colttests(scores1, as.factor(substr(type, 1, 1)))
scores2 <- getSampleMatrix(modules2.robust)
tt2 <- colttests(scores2, as.factor(substr(type, 1, 1)))
sign1 <- which(p.adjust(tt1$p.value, "holm") < 0.05)
sign2 <- which(p.adjust(tt2$p.value, "holm") < 0.05)
@

For the first set of samples \Sexpr{length(sign1)} modules show
significant difference for T-cell and B-cell samples, for the second
(smart) set \Sexpr{length(sign2)} of them.

Let's make some condition plots for the best separating modules from
each set.
<<conditions-plots-dysreg,fig=TRUE,width=10,height=8>>=
color <- ifelse(grepl("T", type), "orange", "darkolivegreen")
layout(cbind(1:2))
cond.plot(modules1.robust, which.min(tt1$p.value), ALL.normed, col=color, 
          main="Best separator, random seeds")
cond.plot(modules2.robust, which.min(tt2$p.value), ALL.normed, col=color,
          main="Best separator, smart seeds")
@

Let's extract the modules that are good separators.
<<extract separators>>=
modules1.TB <- modules1.robust[[sign1]]
modules2.TB <- modules2.robust[[sign2]]
@

\subsection{Enrichment calculations}

We can check our separator modules against the Gene Ontology
categories and the pathways in the KEGG database to find dysregulated
GO categories and/or KEGG pathways.
<<dysreg enrichment>>=
GO.dysreg1 <- ISA.GO(modules1.TB)
GO.dysreg2 <- ISA.GO(modules2.TB)
KEGG.dysreg1 <- ISA.KEGG(modules1.TB)
KEGG.dysreg2 <- ISA.KEGG(modules2.TB)
@

Let's collect the significantly enriched GO categories and KEGG
pathways.
<<dysreg enriched categories>>=
gocats <- unique(unlist(lapply(c(GO.dysreg1, GO.dysreg2), function(x) 
                               unique(unlist(sigCategories(x))))))
keggp  <- unique(unlist(lapply(list(KEGG.dysreg1, KEGG.dysreg2), function(x)
                        unique(unlist(sigCategories(x))))))
library(GO.db)
library(KEGG.db)
sapply(mget(gocats, GOTERM), Term)
mget(keggp, KEGGPATHID2NAME)
@

\subsection{More separator modules}

Of course we can search for modules that can separate the samples of
the ALL subtypes as well, e.g. let's try to find some that
differentiate between type B1 and other B types. 
<<B1 vs others>>=
keep <- grepl("^B[1234]", type)
type.B <- ifelse(type[keep]=="B1", "B1", "Bx")
scores.B1.1 <- scores1[keep,]
scores.B1.2 <- scores2[keep,]
tt.B1.1 <- colttests(scores.B1.1, as.factor(type.B))
tt.B1.2 <- colttests(scores.B1.2, as.factor(type.B))
min(p.adjust(na.omit(tt.B1.1$p.value)))
min(p.adjust(na.omit(tt.B1.2$p.value)))
@

Both module sets seem to have some good separators, let us make some
condition plots.
<<B1-cond-plots,fig=TRUE,width=10,height=8>>=
color <- ifelse(type == "B1", "green", 
                ifelse(grepl("^T", type), "orange", "darkolivegreen"))
layout(cbind(1:2))
cond.plot(modules1.robust, which.min(tt.B1.1$p.value), ALL.normed, col=color, 
          main="Best B1 separator, random seeds")
cond.plot(modules2.robust, which.min(tt.B1.2$p.value), ALL.normed, col=color,
          main="Best B1 separator, smart seeds")
@

From the plot it seems that these two modules are essentially the
same, one is the opposite of the others.
<<check B1 same module>>=
B1.cor <- c(cor(scores1[,which.min(tt.B1.1$p.value)],
                scores2[,which.min(tt.B1.2$p.value)]),
            cor(getFeatureMatrix(modules1.robust, 
                                 mods=which.min(tt.B1.1$p.value)),
                getFeatureMatrix(modules2.robust, 
                                 mods=which.min(tt.B1.2$p.value))))
B1.cor
@

Indeed, they are almost the same, their Pearson correlation is around
\Sexpr{round(B1.cor[1],3)} for the sample scores and
\Sexpr{round(B1.cor[2],3)} for the feature (=gene) scores.
